---
title: "Make the Genoscape"
author: "Eric C. Anderson"
date: "Last Updated: `r Sys.Date()`"
output:
  html_document:
    df_print: paged
    toc: true
  html_notebook:
    toc: true
bibliography: references.bib
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE)
start_time <- Sys.time()
```

# Introduction

The genoscape can be thought of as the bright colors smeared across space that show where different
genetically identifiable groups of birds reside on the breeding grounds.  These genoscapes
are stored as rasters, and transparency is used to indicate how much confidence one has
in the genetic identification of individuals in different areas.  These rasters are made
by interpolating Q-values from a program like STRUCTURE or ADMIXTURE between individuals
that were sampled in space.

The steps to making a genoscape from a matrix of Q-values (i.e., like those
estimated from _structure_) followed here are those that appear on GitHub Pages,
[here](https://eriqande.github.io/make-a-BGP-map/Make-a-BGP-map-Notebook.nb.html).
We reiterate that treatment here, having updated all the links and dependencies
to work in this repository.

This chronicles the steps taken whilst making a genoscape for willow flycatchers.
The order in which we discuss building up this map is different from the order in which
we put layers down to actually make the map. Here, we will start with making the
genoscape, which is actually the part that sits atop the whole map.
We do this, because it doesn't take too long
to plot these features, and using them is a good way to figure out the desired extent of
your map and to decide upon a projection.  Finally, we will talk about the bottom
layer of the map, which is the raster with earthforms and shading from Natural Earth Data.

## Packages Needed

### Non-standard Packages

This work draws on a few functions that I have in packages that I have up on GitHub,
namely:

1. **my fork** of `tess3r`.  Note that you can't use the default version of `tess3r`,
you have to use my fork of it, which has some extra functionality.
2. my package `genoscapeRtools`

You should have already installed those packages as part of the `ruegg-et-al-wifl-genoscape`
depencies.  But, if you have not:
```{r, eval=FALSE}
remotes::install_github("eriqande/TESS3_encho_sen")  # for special version of tess3r
remotes::install_github("eriqande/genoscapeRtools")  # for Eric's genoscapeRtools
```

### Standard Packages

The rest of the packages you need can be downloaded from CRAN.  Again,
you should have already gotten them as dependencies for the repo, but
if not, get `raster`, `sf`, `fields`, `ggspatial`, `downloader`, and `tidyverse`. 
The last one there gets ggplot2
and a number of other packages by Hadley Wickham.

You can get those like this:
```{r, eval=FALSE}
install.packages(c("raster", "sf", "tidyverse", "fields", "ggspatial", "downloader"))
```

### Load the Packages We will work with

```{r, comment=FALSE, warning=FALSE, message=FALSE}
library(raster)  # important to load before tidyverse, otherwise it masks select()
library(tidyverse)
library(sf)
library(ggspatial)
```


# Input Data

## Breeding bird Q values (and colors)

In `004-combine-RAD-and-fludigm-breeders-and-run-STRUCTURE.Rmd` we set up the _structure_
runs to get the $Q$-values.  Of all the runs done there and depicted in Supporting Information
Figure 2, the run that had converged (without any prior information about sampling location)
to the best representation of the spatial arrangement of the samples was $K=7$, replicate
run #4. That run was chosen to make the genoscape.  The Q-values from that run are
stored in this repository in `./stored_results/004/Struct-q-values-k007_Rep004.txt`.

We read them in and work with them.
```{r}
Q_tibble <- read_table2("stored_results/004/Struct-q-values-k007_Rep004.txt", col_names = FALSE)
names(Q_tibble) <- c(
  "id",
  "INW",
  "EST",
  "PNW",
  "SSW",
  "SCC",
  "KER",
  "WMT"
)
```

Here is a quick look at the top of that Q_tibble:
```{r}
head(Q_tibble)
```

Column `id` is the sample name and the rest are ancestry fractions to different clusters
(named with three characters) estimated by STRUCTURE.

Get the colors as well:

```{r}
# we also have colors for those stored in the repo:
source("R/colors.R")

# that has defined this:
cluster_colors
```

First, let us just make a distruct-like plot to remind us of this. We can use
ggplot2 to do that:
```{r}
# first, we want to get the region from the names of the samples, so we can
# print these atop the plot
region_names_and_positions <- Q_tibble %>%
  mutate(
    idx = 1:n(),
    region = str_replace(id, "[rf][0-9]+$", "")
  ) %>%
  group_by(region) %>%
  summarise(
    first = min(idx),
    last = max(idx),
    mid = (min(idx) + max(idx)) / 2
  ) %>%
  arrange(first)

# then we have to pivot the data frame into longer format:
qlong <- Q_tibble %>%
  mutate(idx = 1:n()) %>%
  pivot_longer(
    cols = -c(id, idx),
    names_to = "cluster",
    values_to = "q"
  )

# then plot that dude 
g <- ggplot() +
  geom_col(
    data = qlong,
    mapping = aes(x = idx, y = q, fill = cluster),
    colour = NA
  ) +
  scale_fill_manual(values = cluster_colors) +
  geom_segment(
    data = region_names_and_positions,
    mapping = aes(
      x = first,
      xend = first,
      y = 0,
      yend = 1
    ),
    size = 0.5
  ) +
  geom_text(
    data = region_names_and_positions,
    mapping = aes(
      x = mid,
      y = 1.05,
      label = region
    ),
    angle = 70
  )

g

```





## Breeding Bird Lat-Longs

We also need to know where those individuals were sampled in space. We have that here:
```{r, message=FALSE}
LatLong_tibble <- read_csv("data/appendices/app1-SITable1_Breed_IndvAssigwgenos.csv") %>%
  select(group_af_short_name, Lat, Long) %>%
  rename(
    id = group_af_short_name
  )
head(LatLong_tibble)
```


## Breeding Bird Shapefile

Finally, we need to have a GIS Shapefile that tells us the range of the breeding
birds, so that genoscape can be clipped properly.  We read this shapefile with the
`st_read()` function from package `sf`.
```{r, results='hide'}
breeding_range <- st_read("data/WIFL_RangeMap/WIFLrev.shp")
```

We can plot those polygons to see what they look like:
```{r}
ggplot(breeding_range) +
  geom_sf() +
  theme_bw()
```


# Make the genoscape by kriging and scaling

Within Eric's fork of `tess3r` is a function called `tess3Q_map_rasters`.  It
takes input from the objects we have above, but it takes that input as
matrices rather than data frames, etc. so there is a little finagling to be done.

## First, make sure the lat longs are in the correct order and arrangement

We need to ensure that we have values for birds in the right order (a job for a` left_join`), and we
also have to make it a matrix with Longitude in the first column and Lat in the
second.
```{r}
long_lat_tibble <- Q_tibble %>%
  select(id) %>%
  left_join(LatLong_tibble, by = "id") %>%
  select(Long, Lat)

long_lat_matrix <- long_lat_tibble %>%
  as.matrix()
```

## Then, make a matrix of the Q values

Pull off the names of individuals and make a matrix of it:
```{r}
Q_matrix <- Q_tibble %>%
  select(-id) %>%
  as.matrix()
```

## Interpolate the Q-values by Kriging

For this, we use the above variables in `tess3r::tess3Q_map_rasters()`.  Note the
use of namespace addressing for this function rather than loading the whole `tess3r` package
with the `library()` command.
```{r}
genoscape_brick <- tess3r::tess3Q_map_rasters(
  x = Q_matrix,
  coord = long_lat_matrix,
  map.polygon = breeding_range,
  window = extent(breeding_range)[1:4],
  resolution = c(300,300), # if you want more cells in your raster, set higher
  # this next lines need to to be here, but don't do much...
  col.palette = tess3r::CreatePalette(cluster_colors, length(cluster_colors)),
  method = "map.max",
  interpol = tess3r::FieldsKrigModel(10),
  main = "Ancestry coefficients",
  xlab = "Longitude",
  ylab = "Latitude",
  cex = .4
)

# after that, we need to add names of the clusters back onto this raster brick
names(genoscape_brick) <- names(Q_tibble)[-1]
```

That gives us a raster brick of Q-values associated with each cell in the raster, but
those values are not always constrained between 0 and 1, so we have to massage them
a little bit in the next section.

## Scaling and cleaning the genoscape_brick

For this we use the function `genoscapeRtools::qprob_rando_raster()`.  This takes the raster
brick that comes out of tess3Q_map_rasters() and does some rescaling and (maybe) some random sampling
to return a raster of colors that I hope will do a reliable job of representing (in some way) predicted
assignment accuracy over space.
See `?genoscapeRtools::qprob_rando_raster` to learn about the scaling options, etc. 

This will squash the raster brick down to a single
RGBA (i.e., four channels, red, green, blue and alpha) raster brick.
```{r}
genoscape_rgba <- genoscapeRtools::qprob_rando_raster(
  TRB = genoscape_brick,
  cols = cluster_colors,
  alpha_scale = 2.0,
  abs_thresh = 0.0,
  alpha_exp = 1.55,
  alpha_chop_max = 230
)

# at this point, we must be explicit about adding a projection to the raster.
# This adds the info for a regular lat-long projection
crs(genoscape_rgba) <- "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"



```

We can easily plot this with the function `layer_spatial` from the `ggspatial` package:
```{r}
ggplot() +
  ggspatial::layer_spatial(genoscape_rgba) +
  theme_bw() +
  coord_sf()
```


# Sample Location Membership

From this plot it is clear which sampling location belong to which regions/clusters/colors
of the genoscape.  Breeding birds were placed into such colors/clusters according to their
spatial location (and not by their individual Structure Q-values, etc.)  We do that
here:
```{r}
source("R/assign_points_to_raster_regions.R")

# let's get the ids and the lat-longs of all these birds
# again, so we can check what we have in the appendix
LLG_tibble <- read_csv("data/appendices/app1-SITable1_Breed_IndvAssigwgenos.csv") %>%
  select(group_af_short_name, Lat, Long, repunit, Stage) %>%
  rename(
    id = group_af_short_name
  )

# now, we find which section of the genoscape each point is in
LLG_with_spatial <- assign_points_to_raster_regions(
  genoscape_brick,
  LLG_tibble
)
```

The results here have been recorded in the data supplement: `app1-SITable1_Breed_IndvAssigwgenos.csv` in the `repunit` column.

We show here that the `repunit` column and the `spatial_group`
(just computed) are
same for all birds except for the samples from Bishop, CA:
```{r}
LLG_with_spatial %>%
  filter(repunit != spatial_group)
```
This one sampling location was indeterminate in the genoscape, lying
at the intersection of four different clusters/colors.  Further, the sample
was populated with individuals that showed genetic affiliation to a number
of different groups, suggesting that some may have been migrants.  Furthermore,
the population in Bishop is very small and thus would not be expected
to contribute birds to the wintering samples.  Therefore, in
`app1-SITable1_Breed_IndvAssigwgenos.csv`, their repunit is designated
as `BIS`, but they will be removed from the reference data set for
identification of wintering birds.

The following plot shows the location of the Bishop samples highlighted
by a large, black circle.  Note that it does not show clear affiliation
with any of the other regions, unlike the other sampling points.
```{r}
ggplot() +
  layer_spatial(genoscape_rgba) +
  geom_spatial_point(
    data = long_lat_tibble, 
    mapping = aes(x = Long, y = Lat), size = 0.2
  ) +
  annotate(
    "point", 
    x = -118.4800,
    y = 37.4083,
    size = 9,
    shape = 21,
    fill = NA
  ) +
  theme_bw() +
  coord_sf()
```



# Other map elements


Since we will be using the Natural Earth Data Set for the raster background of our map,
we will also use the Natural Earth lines and polygons.  The Natural Earth data set is
an amazing, open-source resource for making beautiful maps.  Check it out at
[naturalearthdata.com](https://www.naturalearthdata.com/).

## Downloading the shapefiles for Coastlines, Countries, States, and Provinces

For coastlines, countries, and states/provinces, you will need to download
three different shape files.  We will be working with the highest resolution
Natural Earth data sets which are the `10m` versions. 
According to the dependencies, (see the README), you should have already downloaded
these and unzipped them
to a directory within the top level of the RStudio project called `geo-spatial`.

## Plot the Coastlines and estimate where we want to clip

First read it in:
```{r}
coastlines <- st_read("geo-spatial/ne_10m_coastline/ne_10m_coastline.shp")
```
Now, let's just crop out the part that we want.  This is somewhat key:
right here we will define the extent in lat-long of the region that
we want to plot:
```{r}
# note! it is important to put the elements of domain in this
# order, because the function raster::extent() is expecting things
# to be in this order, and doesn't parse the names of the vectors
# the way sf::st_crop() does.
domain <- c(
  xmin = -135,
  xmax = -60,
  ymin = 22,
  ymax = 60
)

coast_cropped <- st_crop(coastlines, domain)

```

Then plot the cropped part:
```{r}
ggplot(coast_cropped) +
  geom_sf() +
  coord_sf()
```

## All the lines, plus the genoscape

OK, that was pretty reasonable.  Now crop all the lines to `domain` and plot
them, and put the genoscape on top of that, too.
```{r}
countries_cropped <-  st_read("geo-spatial/ne_10m_admin_0_boundary_lines_land/ne_10m_admin_0_boundary_lines_land.shp") %>%
  st_crop(domain)
states_cropped <- st_read("geo-spatial/ne_10m_admin_1_states_provinces_lines/ne_10m_admin_1_states_provinces_lines.shp") %>%
  st_crop(domain)
```

Now, plot it all.  Notice we do it by just adding layers.
```{r}
mapg <- ggplot() +
  geom_sf(data = coast_cropped) +
  geom_sf(data = countries_cropped, fill = NA) +
  geom_sf(data = states_cropped, fill = NA) +
  ggspatial::layer_spatial(genoscape_rgba) +
  theme_bw()

# now plot it under default lat-long projection
mapg +
  coord_sf()
```

That is looking like it should, but the projection is pretty ugly.  In the next section
we will consider projection.

## Projecting

The new version of `ggplot2` makes it super easy to project everything on the fly
by passing the projection to `coord_sf()`.   For things in North America,
it seems that a Lambert conic projection does a nice job of keeping British Columbia
and Alaska from looking way too big. We can define such a projection with a string, like
this:
```{r}
lamproj <- "+proj=lcc +lat_1=20 +lat_2=60 +lat_0=40 +lon_0=-100 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs"
```

Now, let's see how our developing map looks under such a projection.  We just define `lamproj` as
a coordinate reference system and pass it in to `coord_sf()`:
```{r}
mapg +
  coord_sf(crs = st_crs(lamproj))
```

I would call that decidedly better.

## The pretty-map background

Now, all that remains is to put all of this on top of a nicely tinted map that
shows landforms and things.  Note that once we have such a layer under the rest of this
stuff, we will probably omit the coastline layer, since it gets a little dark where the
coastline is highly dissected.

### Read in the raster and crop it

The raster data should already have been downloaded and placed in the appropriate
location by dealing with the dependencies in the README.

Reading a large raster in with the raster package does not take up much
memory, because it leaves it on disk.  So we will open a connection to
the big, massive raster using the `brick` function from the `raster` package
and then crop it:
```{r}
hypso <- brick("geo-spatial/HYP_HR_SR_OB_DR/HYP_HR_SR_OB_DR.tif")
hypso_cropped <- crop(hypso, extent(domain))
```

Now, we just add `hypso_cropped` in with `ggspatial::spatial_layer()` below all of our
other layers (dropping the coastlines), and we might make country and state lines thinner (and different sizes):
```{r}
big_one <- ggplot() +
  ggspatial::layer_spatial(hypso_cropped) +
  geom_sf(data = countries_cropped, fill = NA, size = 0.15) +
  geom_sf(data = states_cropped, fill = NA, size = 0.1) +
  ggspatial::layer_spatial(genoscape_rgba) +
  theme_bw() +
  coord_sf(crs = st_crs(lamproj))

big_one
```

Seeing that sort of small on the screen does not really do justice to it.  You can `ggsave` it
in whatever size is desired.  For whatever the final product is going to be, you will
want to mess with the line thickness on those country and state borders...

## Clipping a rectangle out of that

While that looks nice, most people might prefer to have the
background part there in a rectangle, rather than a conic surface.  We can
do that by setting the xlim and ylim to `coord_sf()`.  But, note that
we have to do that in terms of the units that the Lambert projection is referring to,
not just in simple lat-long coordinates (i.e. we have to project those too!)

So, imagine that we want to chop out a rectangle from Haida Gwai, off the coast of
British Columbia, straight down and straight across to the right (on the image).
And we would want the bottom and right sides determined by a point in the bottom
right that has the y-value of the tip of Florida, and an x-value which is essentially
at St. John, New Brunswick.  Then, we need to determine the x and y coordinates of those
points under our Lambert conformal conic projection.  We do that by making a simple features
tibble of the lat-longs for those points and then projecting it.

```{r}
# Here are lat longs for those points
pts <- tribble(
  ~place, ~long, ~lat,
  "HaidaGwai", -132.3143526, 53.7298050,
  "FloridaTip", -80.8032237, 25.0909780,
  "St.Johns",  -65.943453, 45.291222
)

# here we turn that into an sf object
pts_sf <- st_as_sf(pts, coords = c("long", "lat"),
                   crs = 4326)  # 4326 is shorthand for "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"

# and now we transform it to lambert
pts_lambert <- st_transform(pts_sf, crs = lamproj)

# when we print it in a notebook, we can see that the geometry is a list
# column, but can't see the values:
pts_lambert
```

So, for you to be able to see the values we can just print it like this:
```{r}
# just note the values:
st_as_text(pts_lambert$geometry)
```

That means we want xlim = c(-2010114, 2452877) and ylim = c(-1366911, 1821869).
Let's try that:
```{r}
rectangled <- ggplot() +
  ggspatial::layer_spatial(hypso_cropped) +
  geom_sf(data = countries_cropped, fill = NA, size = 0.15) +
  geom_sf(data = states_cropped, fill = NA, size = 0.1) +
  ggspatial::layer_spatial(genoscape_rgba) +
  theme_bw() +
  coord_sf(
    crs = st_crs(lamproj),
    xlim = c(-2010114, 2452877),
    ylim = c(-1366911, 1821869),
    expand = FALSE)  # expand = FALSE means put the axes right at the xlim and ylim

rectangled
```

Okay, that does it.  In reality, you would probably want to expand the original `domain`
to be a bigger chunk of the earth, so that when you cut stuff off there was more left...


# A final note

If you want to add more elements to the figure you can do so by giving them
a lat-long and making a simple features tibble out of it and then putting them
there using geom_sf().  It all fits nicely
into the ggplot framework.

Also, the final genoscape figure in the paper included the wintering birds
as well. For reasons of computing time, we aren't going to show the exact
code for putting that together---it takes a long time to render.  But,
it should be obvious from the above how it was done.

# Running Time

Running the code and rendering this notebook required approximately this much time
on a Mac laptop of middling speed, listed in `HH:MM:SS`.
```{r}
td <- Sys.time() - start_time
tdf <- hms::as_hms(ceiling(hms::as_hms(td)))
dir.create("stored_run_times", showWarnings = FALSE, recursive = TRUE)
write_rds(tdf, "stored_run_times/006.rds")
tdf
``` 

# Session Info

```{r}
sessioninfo::session_info()
```

