---
title: "Ruegg et. al., WIFL Genoscape, Data and Code Repository"
author: "Kristen Ruegg, Eric C. Anderson, Marius Somvielle, Rachael A. Bay, Christen Bossu, Mary Whitfield, Eben H. Paxton, Thomas B. Smith"
output:
  html_document:
    toc: true
    toc_depth: 3
    df_print: paged
  github_document:
    toc: true
    toc_depth: 3
bibliography: references.bib
---

**Last Updated:** `r Sys.Date()`


```{r setup, include=FALSE, message=FALSE}
library(tidyverse)
library(hms)

knitr::opts_chunk$set(echo = TRUE)

# use this for the the README.md
GPbase <- "https://eriqande.github.io/ruegg-et-al-wifl-genoscape/"

# use this for making the README.html that we will hard link to docs/index.html
Dbase <- "./"

# choose the one that you want here depending on if you are on the
# GitHub README or are making an index.html for the docs directory
hdir <- GPbase

make_link <- function(Rmd, base = hdir) {
  nb <- stringr::str_replace(Rmd, "Rmd$", "html")
  sprintf("%s\n\n(_Compiled RMarkdown HTML document on GitHub Pages:_ [%s](%s%s))\n\n", Rmd, nb, base, nb)
}
```

# Overview

This repository includes code, data, and some intermediate results to
reproduce the results in Ruegg et al. ("Seasonal niche breadth predicts
population declines in a long-distance migratory bird").
You can get the whole thing by cloning or downloading the repo from
[https://github.com/eriqande/ruegg-et-al-wifl-genoscape](https://github.com/eriqande/ruegg-et-al-wifl-genoscape).

If you are viewing this as the README of a GitHub repository, note that
you can read it in a somewhat friendlier format (i.e., with a table
of contents, etc.) on GitHub pages at:
[https://eriqande.github.io/ruegg-et-al-wifl-genoscape/](https://eriqande.github.io/ruegg-et-al-wifl-genoscape/)


RAD sequening and the production of BAMs and VCFs was described and documented in
@ruegg2018ecological. The work described in this repository uses the products from that work

The various steps in our analyses are described in a series of RMarkdown documents with a
fair bit of explanation/description of the procedures.  The few analyses that used
cluster computing resources are described and documented, but results from those analyses
are included in `stored_results` so that it is not necessary to have access to a cluster
to reproduce the results herein. 

Running some of the RMarkdown documents requires
a proper shell.  These were initially developed and run on a Mac.

Rmarkdown documents and scripts in the 000 series were developed primarily by Eric C. Anderson.
Rmarkdown documents and scripts in the 100 series were developed primarily by Marius Somveille.



All the RMarkdown documents or scripts can be evaluated en masse by
sourcing the R script `render-numbered-Rmds.R`.  On a fairly old mac laptop the run times
for each are as follows:
```{r, echo=FALSE}
if(!file.exists("stored_run_times")) {
  message("You must run all the Rmarkdown documents to store the run times...")
} else {
  rtfs <- dir("stored_run_times", full.names = TRUE)
  time_tib <- tibble(
    rts = lapply(rtfs, function(x) read_rds(x)),
    tfile = basename(rtfs)
  ) %>%
    mutate(`Running time in HH:MM:SS` = map_chr(rts, format))
  
  rmds <- dir(pattern = "[0-9][0-9][0-9]-.*\\.Rmd")
  file_tib <- tibble(File = rmds) %>%
    mutate(tfile = str_c(str_match(rmds, "([0-9][0-9][0-9])-.*\\.Rmd")[, 2], ".rds"))
  fft <- file_tib %>%
    left_join(time_tib, by = "tfile") %>%
    select(File, `Running time in HH:MM:SS`)
  fft
}
```


# Preliminaries and Dependencies

Below is a description of the needed dependencies.  A script to install all of
them on one of our test clusters is at `000-00-prepare-dependencies.sh` (NOT YET INCLUDED), 
but it may likely need some tweaking on your system.

## Unix Programs

The following programs must be installed and available in the PATH. Versions
used on Eric's laptop appear in parentheses. (NOT COMPLETE YET)

- `bgzip`:
- `samtools`:
- `tabix`:


## Genomic resources

Create a directory in the top level of the repository, download the willow flycatcher
then modify the sequence names to be
consistent with the nomenclature used in this work and index the genome (that
nomenclature is already in in the FASTA file, but just not as the sequence name
so, we need to only shift some words around on each sequence name line).
```sh
mkdir genome
cd genome/
wget ftp://ftp.ncbi.nlm.nih.gov/genomes/genbank/vertebrate_other/Empidonax_traillii/latest_assembly_versions/GCA_003031625.1_ASM303162v1/GCA_003031625.1_ASM303162v1_genomic.fna.gz
ftp://ftp.ncbi.nlm.nih.gov/genomes/genbank/vertebrate_other/Empidonax_traillii/latest_assembly_versions/GCA_003031625.1_ASM303162v1/GCA_003031625.1_ASM303162v1_assembly_report.txt

# now, change the sequence names and make sure that the DNA bases
# are all listed in UPPERCASE.
gzcat GCA_003031625.1_ASM303162v1_genomic.fna.gz | awk '
  /^>/ {
    tmp = substr($1, 2)
    newname = $6
    sub(/,/, "", newname)
    sub(/:/, "|", newname)
    $1 = ">" newname;
    $6 = tmp ",";
    print
    next
  }
  {print toupper($0)}
' > wifl-genome.fna
samtools faidx wifl-genome.fna
```

## Geospatial data

Make a directory in the top level of the repository called
`geo-spatial` and then download some large files from Natural Earth Data
to there:

- Natural Earth II with Shaded Relief, Water, and Drainages raster [https://www.naturalearthdata.com/http//www.naturalearthdata.com/download/10m/raster/NE2_HR_LC_SR_W_DR.zip](https://www.naturalearthdata.com/http//www.naturalearthdata.com/download/10m/raster/NE2_HR_LC_SR_W_DR.zip).  Put or symlink the resulting directory, `NE2_HR_LC_SR_W_DR` into `geo-spatial`.
- 10m-cultural-vectors, Admin 1 – States, Provinces, Download boundary lines: [https://www.naturalearthdata.com/http//www.naturalearthdata.com/download/10m/cultural/ne_10m_admin_1_states_provinces_lines.zip](https://www.naturalearthdata.com/http//www.naturalearthdata.com/download/10m/cultural/ne_10m_admin_1_states_provinces_lines.zip). Put or symlink the resulting directory, `ne_10m_admin_1_states_provinces_lines` into `geo-spatial`.
- Finally, get the coastlines: [https://www.naturalearthdata.com/http//www.naturalearthdata.com/download/10m/physical/ne_10m_coastline.zip](https://www.naturalearthdata.com/http//www.naturalearthdata.com/download/10m/physical/ne_10m_coastline.zip), and put the resulting folder, `ne_10m_coastline` into `geo-spatial`.




## Java jars

The following Java-based programs must be downloaded, and the paths to their
associated Jar files must be listed appropriately in the file `script/java-jar-paths.R`.
(NO JAVA JARS NEEDED AT THIS TIME...)


## R Packages

Packages must be downloaded from CRAN, BioConductor, and GitHub.  The following
code, which is in `R/install_packages_etc.R`, will download and install the necessary
packages:

```r
`r paste(readLines("R/install_packages_etc.R"), collapse = "\n")`
```

# RMarkdown Documents

The following RMarkdown documents should be evaluated in order.
The script `render-numbered-Rmds.R` (NOT WRITTEN YET) will do that
when run in the top level
of this repository.
Some RMarkdown documents rely on
outputs from previous ones.  Some of these RMarkdown documents include
calls to Unix utilities, so might not run on non-Unix or non-Linux architectures.

Outputs (figures, tables, R data objects, etc) from each RMarkdown document
are written to the `outputs/XXX` directories.  Intermediate files
written during evaluation of each RMarkdown document are written
to the `intermediates/XXX` directories (THOUGH I MIGHT NOT USE intermediates AT ALL). 
To facilitate working between
the cluster and a desktop/laptop, some outputs are written to
the `stored_results/XXX` directories which are version controlled and
included in this repo.


Thumbnails of the figures and tables generated by each RMarkdown document appear
below (NOT YET).  Additionally, at the top of each section is a link to the
compiled (HTML-version) of the RMarkdown document on GitHub Pages.


## `r make_link("001-rad-seq-data-summaries-and-pca.Rmd")`

Summarizing the RAD seq data, filtering out obvious paralogs and monomorphic sites,
then producing a PCA plot and estimates of pairwise $F_\mathrm{ST}$. 

## `r make_link("002-select-snps-for-assays-from-rad.Rmd")`

All the steps we went through to rank SNPs for ability to resolve different
populations/subspecies, and then the workflow for designing Fluidigm assays from them.

# Literature Cited

